Food-web structure, stochasticity and response diversity drive temporal community stability in complex communities

### 2024-04-10 paper/manuscript.Rmd
Reframe introduction: shorter and more general

---
title: Food-web structure, stochasticity and response diversity drive temporal community stability in complex communities
author: "Alain Danet (1), Sonia Kéfi (2,3), Thomas F Johnson (1), Andrew P. Beckerman (1)"
date: ""
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    spacing: double
    number_sections: false
always_allow_html: true
bibliography: references.bib
link-citations: true
csl: ecology-letters.csl
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
---

@@

@@

Empirical and theoretical evidence suggests environmental changes will
dramatically destabilise ecological communities. However, this evidence is
largely based on oversimplified communities, which lack the complexity of
natural communities. We characterise the dynamics of complex food-webs under a
stochastic food-web model, and investigate how response diversity,
stochasticity, and food-web structure drive temporal stability. We found that
higher average trophic level increased community stability by increasing
population stability, at the expense of asynchrony. Species richness had little
effect on population stability, but a large one on asynchrony. Interaction
strength and connectance had little effects on temporal stability overall, but
interacted strongly with response diversity to determine stability-richness
relationships. Absence of response diversity all leaded to negative
stability-richness relationships, but high response diversity resulted in high
prevalence of positive stability-richness relationships. Our study reveals that
response diversity and food-web structure drive temporal stability overall, and
together determine the sign of the stability-richness relationship.

@@

Global environmental change constitutes a suite of major threats to biodiversity
and ecosystem functioning [@lee_ipcc_2023; @simmons_refocusing_2021]. A central
feature of these threats is that they are predicted to dramatically increase
variability in the environment that species face [@lee_ipcc_2023] with potential
impacts on the stability of communities and associated impacts on biodiversity
and functioning in these communities. Assessing the impacts of this increased
variability on ecological communities is challenging for several reasons.

First, the stability of communities is often evaluated using the temporal
variability of a metric of interest (e.g. abundance or biomass), because other
stability metrics such as resilience and resistance are not always applicable,
as most natural communities face persistent perturbations [@arnoldi_resilience_2016].
Understanding this temporal community variability (hereafter referred to as
community stability), measured as the inverse of the coefficient of variation
[@tilman_biodiversity_2006; @roscher_identifying_2011;
@thibaut_understanding_2013; @zhao_biodiversity_2022], requires scaling down the
focus from communities to populations.
Community stability can be mathematically partitioned into average population
stability and asynchrony [@loreau_species_2008;
@thibaut_understanding_2013; @zhao_biodiversity_2022] - asynchrony measuring how
much each species fluctuations’ are compensating for each other at the community
level. Population stability and asynchrony both depend on the number of species
and their dominance [@loreau_species_2008; @thibaut_understanding_2013;
@zhao_biodiversity_2022], on the interactions among species [@thebault_trophic_2005;
@thebault_relationship_2006; @loreau_species_2008; @danet_species_2021;
@eschenbrenner_diversity_2023] and on the way species respond to environmental
perturbations [@ives_stability_2000; so-called response diversity,
@elmqvist_response_2003; @ross_how_2023]. Therefore, accounting for the effects
of community structure on population stability and asynchrony is crucial to
improve our understanding of community stability in facing environmental change.

Average population stability decreases with environmental stochasticity, such as the
intensity and the frequency of perturbations [@loreau_species_2008]. Theorerical
and empiral evidence suggests that species richness can decrease
[@tilman_biodiversity_2006; @loreau_species_2008], increase
[@houlahan_negative_2018], or have little effects on population stability. This
effect of species richness on population stability depends on the evenness of
the community since dominant species contribute more to the average of the
community than subordinate species. Species interactions also have tremendous
effects on population stability, such as strong interaction strength are
generally destabilising [@mccann_weak_1998; @mccann_reevaluating_1997;
@thebault_trophic_2005]. The shape of food-webs can also modulate population
stability, such as higher trophic level was generally found to be stabilising
[@shanafelt_stability_2018; @danet_species_2021; @eschenbrenner_diversity_2023],
as top-down and bottom-up controls can increase population stability
[@shanafelt_stability_2018].

Asynchrony is mostly determined by the characteristics of the network. Response
diversity captures how species’ differences in their response to the
environmental variation can lead to asynchronous dynamics if their responses are
different but synchronous dynamics if their responses are similar. Theory and
empirical data emerging around response diversity suggest that it is a “core
ecological mechanism underpinning many of the statistical explanations of the
diversity–stability relationships” [@ives_stability_2000;
@elmqvist_response_2003; @ross_how_2023]. In turn, the network structural
properties and strength of species interactions can also modulate asynchrony
through compensatory dynamics such as a strong asymmetry of consumer preference
can generate asynchrony in resource fluctuations [@mccann_weak_1998;
@vasseur_environmental_2007], and a generalist consumer can in turn generate
synchrony among its prey [@raimondo_interspecific_2004]. Competition can also
have synchronising effects in strong species growth rates regime
[@loreau_species_2008]. Finally, portfolio and compensatory effects can also be
affected by the population evenness such as fluctuations by dominant species
will hardly be compensated by a subordinate species
[@thibaut_understanding_2013; @zhao_biodiversity_2022]. Improving our
understanding of community stability facing perturbations therefore requires
articulating the effects of response diversity, species interactions, species
richness and evenness on the different components of stability. Predicting the
effects of species richness on community stability thus requires understanding
the relationship between species richness, population stability and asynchrony.

Empirical evidence suggest by increasing asynchrony, species richness
drives positive relationships between community stability and species richness
[@tilman_biodiversity_2006; @isbell_biodiversity_2009]. In contrast population stability was
found to either decrease, increase, or not change with species richness
[@houlahan_negative_2018; @tilman_biodiversity_2006; @roscher_identifying_2011].
We note, however, that evidence for both asynchrony-richness and population
stability-richness is derived from data on species assemblages.  These data thus
do not account for food-web structure, which can have significant impact on
population stability and asynchrony [but see @danet_species_2021]. Thus, ecology
lacks a body of theory of temporal stability integrating across structure and
response diversity frameworks and reflecting species rich communities. Here, we
integrate these two perspectives on diversity- stability to evaluate the
combined effects of community structure and response diversity on temporal
stability in species rich, complex communities characterised by competition and
predation.

The ability to do this is supported by a key statistical advance: it is now
possible to partition asynchrony into two component sources: the portfolio
and the compensatory dynamics effects [@zhao_biodiversity_2022]. Portfolio
effects stem from the statistical averaging of independent population
fluctuations [@doak_statistical_1998; @zhao_biodiversity_2022], thereby
increasing asynchrony. Compensatory dynamics in turn capture population
fluctuations that are more synchronous or asynchronous than expected for
independent fluctuations, and can stem from species interactions and species
responses to perturbations [@zhao_biodiversity_2022]. This allows an
unprecedented opportunity to obtain mechanistic insight about the effects of
stressors and food-web structure on stability in complex, species rich
communities.

Our primary objective is to derive theory that predicts how the components of
community stability, that can differentially drive patterns of stability,
deliver novel insights into the source and patterns of diversity-stability
relationships in complex communities facing environmental stochasticity, a
necessary ingredient to generate species fluctuations. Our approach delivers
revised insight into diversity-stability relationships in species rich
communities when both community structure per se, and species responses can
vary. In the following sections, we begin by introducing the multi-species model
and the features of the model that allow us to integrate across variation in
community structure and response diversity. We then introduce assumptions we
used to implement environmental stochasticity, response diversity, variation in
community structure and variation in the distribution of interaction types and
strengths. Next, we define our in-silico experimental design that specifically
aligns with three core questions and associated analyses. Considering four
drivers, namely food-web structure, interaction strength, environmental
stochasticity and response diversity, we investigated (1) What are their direct
effects on population stability, portfolio and compensatory effects ? (2) What
are their total effects on community stability, population stability and
asynchrony? Finally, merging the two parallel bodies of work, we asked (3) How
do they interact to drive the sign of the relation between community stability
and species richness?

>>>>

---
title: Food-web structure, stochasticity and response diversity drive temporal community stability in complex communities
author: "Alain Danet (1), Sonia Kéfi (2,3), Thomas F Johnson (1), Andrew P. Beckerman (1)"
date: ""
output:
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
  bookdown::pdf_document2:
    latex_engine: xelatex
    keep_tex: true
    spacing: double
    number_sections: false
always_allow_html: true
bibliography: references.bib
link-citations: true
#csl: nature.csl
csl: ecology-letters.csl
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
---

@@

- Putative titles:
  - Community stability is driven by biodiversity and response diversity in
    complex stochastic food-webs
  - Response diversity and community structure drive community stability in
    complex food-webs

- Two strong results:
  - Community stability raises primary from richness and average trophic level
    but average interaction strength and connectance have little effects
  - Response diversity and community structure interact strongly to determine
    the sign of the relationship between community stability and species
    richness

@@

[NOT CHANGED] Empirical and theoretical evidence suggests environmental changes will
dramatically destabilise ecological communities. However, this evidence is
largely based on oversimplified communities, which lack the complexity of
natural communities. We characterise the dynamics of complex food-webs under a
stochastic food-web model, and investigate how response diversity,
stochasticity, and food-web structure drive temporal stability. We found that
higher average trophic level increased community stability by increasing
population stability, at the expense of asynchrony. Species richness had little
effect on population stability, but a large one on asynchrony. Interaction
strength and connectance had little effects on temporal stability overall, but
interacted strongly with response diversity to determine stability-richness
relationships. Absence of response diversity all leaded to negative
stability-richness relationships, but high response diversity resulted in high
prevalence of positive stability-richness relationships. Our study reveals that
response diversity and food-web structure drive temporal stability overall, and
together determine the sign of the stability-richness relationship.

@@

Ever-growing anthropogenic pressures have resulted in alteration of biodiversity
[@newbold_global_2015; @services_summary_2019], community changes
[@dornelas_assemblage_2014], and climate change will result in dramatic
environmental fluctuations [@lee_ipcc_2023]. How ecological communities will
respond to those threats is then a pressing, but challenging question
[@simmons_refocusing_2021; @yang_predictability_2019]. The temporal stability of
ecological communities is an emerging property that indeed results of the
complex interplay between anthropogenic pressures [@olivier_urbanization_2020;
@hautier_anthropogenic_2015; @yang_predictability_2019], community size
[@tilman_biodiversity_2006; @isbell_biodiversity_2009;
@olivier_urbanization_2020; @pennekamp_biodiversity_2018] and community
structure [@thebault_trophic_2005; @thebault_relationship_2006;
@thibaut_understanding_2013; @danet_species_2021;
@eschenbrenner_diversity_2023].

Understanding emerging temporal community stability (hereafter referred to as
community stability) requires scaling down the focus to population scale
[@doak_statistical_1998; @thibaut_understanding_2013; @zhao_biodiversity_2022].
Community stability can be partitioned into average population
stability and asynchrony [@loreau_species_2008; @thibaut_understanding_2013;
@zhao_biodiversity_2022], asynchrony measuring how much population
fluctuations’ are compensating at the community level. The question of how
global change will respectively affect population stability and asynchrony is
still open.

Asynchrony among species is the central pattern that drive the positive
community stability-diversity relation documented across many species assemblages
[@tilman_biodiversity_2006; @olivier_urbanization_2020;
@pennekamp_biodiversity_2018]. Asynchrony does not emerge by species richness
itself, but by its interplay with response diversity of species to perturbations
[@elmqvist_response_2003; @loreau_species_2008; @ives_stability_2000;
@ross_how_2023], and compensatory dynamics raising from species interactions.
Asynchrony can then stem from statistical averaging of independent species
fluctuations ["portfolio effects", @doak_statistical_1998;
@tilman_ecological_1999] or compensatory dynamics resulting from response
diversity and species interactions [@zhao_biodiversity_2022]. Among other
reasons, it might explain why reports of neutral or negative stability-richness
relationships are also common in empirical settings [@houlahan_negative_2018].
However, most of our knowledge comes from simplified community
structure, i.e. which does not account for the multiple trophic
levels constituting ecological communities [but see @danet_species_2021;
@thebault_trophic_2005; @ives_stability_2007; @eschenbrenner_diversity_2023],
thereby limiting our understanding of the drivers of both community stability
and stability-richness relationships in more complex communities.

The structure of community also shapes communities
[@macarthur_fluctuations_1955; @may_will_1972; @angelis_stability_1975;
@mccann_weak_1998; @vasseur_environmental_2007; @thebault_trophic_2005], such as
the strength [@may_will_1972; @mccann_weak_1998; @thebault_trophic_2005] and
distribution [@may_will_1972; @angelis_stability_1975; @mccann_weak_1998;
@mccann_reevaluating_1997] of species interactions determine population
stability and stability-richness relationship [two trophic level systems,
@thebault_trophic_2005; @ives_stability_2007], while the number of trophic
levels determine population stability [@shanafelt_stability_2018]. Community
structure also interact with response diversity to population stability in
diamond shaped food-webs [@ripa_food_2003; @vasseur_environmental_2007;
@ruokolainen_environmental_2013]. However, the work on community structure
lacked the complexity in terms of species richness or trophic levels, and
focused more on population stability than on asynchrony.

Then, the two body of work focusing on the effects of (1) species richness or
the effects of (2) community structure on community stability have been
developed in isolation. While (1) highlighted the importance of asynchrony
stemming response diversity ("insurance hypothesis"), (2) highlighted the
importance of population stability arising from species interactions (top-down
and bottom-up controls). Two keys advances allow to overcome that limitation:
the development of the Bioenergenetic model that enable to model arbitrary
complex communities [@yodzis_body_1992; @brose_allometric_2006] and the
development of statistical tools that enables to partition community stability
into population stability and asynchrony [@thibaut_understanding_2013;
@zhao_biodiversity_2022].

Here, we investigated the effects of species richness and community structure on
community stability and on stability-richness relationship in complex stochastic
food-web. To do so, we developed an extension of the Bioenergetic model
[@brose_allometric_2006; @lajaaiti_ecologicalnetworksdynamicsjl_2024] which
includes environmental stochasticity and response diversity
[@vasseur_environmental_2007]. We modelled food-web varying in size (species
richness XX) and structure (Connectance XX, Predator-Prey Mass Ratio: XX) to
create a diversity of community structure. We additionally varied the
variability of environmental stochasticity (standard deviation of mortality
rates) and response diversity ($1-\rho$, $\rho$ being the correlation among
species mortality rates). We performed XXXXX simulations over a minimum 2000
timesteps (Table S1) and collected the 500 last timesteps to compute community
stability and structure. Using a structural equation model, we deciphered the
ways in which stressors (environmental stochasticity, decreasing response
diversity) and community structure (species richness, average interaction
strength, average trophic level, connectance) determine community stability. In
a second time, we used a linear model to investigate how community structure
interact with response diversity to determine the sign of the stability-richness
relationship.
